---
title: "The corpus"
---

```{r}
#| include: false
source("R/setup.R")
```

**Real files only.** Toy JSON is hard in ways nobody suffers from, and a tool
tuned against invented documents is tuned against invented problems. Every entry
is something that came off a real API, a real export or a real tool.

```{r}
#| echo: false
d <- sort(list.dirs(file.path(ROOT, "corpus"), recursive = FALSE))
info <- data.frame(
  entry = basename(d),
  size  = sapply(d, function(p) {
    f <- file.path(p, "source.json")
    if (file.exists(f)) format(structure(file.size(f), class = "object_size"),
                               units = "auto", digits = 0)
    else "fetch.sh"
  }),
  row.names = NULL, check.names = FALSE)
knitr::kable(info, col.names = c("entry", "source"))
```

An entry is finished only when all of these are true, and anything less is a
file somebody downloaded:

- `source.json` is committed, or a `fetch.sh` is, with the file described as it
  was on the day.
- Its notes carry provenance with a date, a **measured** number in every grading
  row that can carry one, and a section saying what the file **disconfirmed**.
- One attempt file per tool, in both languages, each with its scoring header
  filled in.
- An entry in the findings saying what the file changed.

## The disconfirmation rule is the point

That third bullet in the notes — *what this file disconfirmed* — is why the
corpus is worth having.

**The first file was chosen as a polymorphism specimen and has none.** That is
worth more than a file which confirmed what somebody already believed, and it is
only worth anything because the expectation was written down first.

Each of the first three files contradicted something. A later one was chosen
specifically as the case where an existing describer should finally be right,
and it dropped the field held by the majority of records. Another was chosen
because two spellings of one field are its whole point, and a database engine
silently collided them and returned the other population's data.

## Why a corpus and not a benchmark

A benchmark asks which tool is fastest at a task everyone agrees on. There is no
agreed task here — that is the problem. The corpus exists to find out whether
there *is* a capturable pattern underneath, by asking the same questions of
documents that are as unlike each other as could be found.

Gaps are tracked as deliberately as entries. The one that matters most is
`scale`: nothing here has ever strained memory, so the axis has never had a
chance to separate anything.
