---
title: "What is in here?"
---

```{r}
#| include: false
source("R/setup.R")
```

The hypothesis this project is testing, written down before anything was
measured:

> Every existing tool makes you state **paths**, from which the row shape
> emerges as a side effect. That is backwards. The syntax should make you state
> **what a row is**, and find the paths itself.

The first version of that claim went on to say *"and no tool helps you explore"*,
which is measurably false. At least six try. They do not fail to describe. They
fail like this:

> **Every existing describer's output is proportional to the data. What is
> needed is output proportional to the structure.**

## The claim as a slope, not a ratio

*"61% the size of the file it describes"* is one document and one point, and a
single ratio cannot distinguish the two hypotheses. So describe a **growing
prefix** of one document and watch what the ratio does:

- proportional to the **data** → the ratio stays **flat**
- proportional to the **structure** → the ratio **falls**, because the structure
  stops growing while the document does not

That is a measurement that could have come out the other way, and it is the
first form of this claim that could. On the corpus's most keys-as-data file the
existing describer is flat across a 9× growth while the probe's input grows 52×
and its answer grows 1.9×.

## The fold

The operation that does it is one idea. **Objects that are siblings of the same
kind are described once, not once each.**

Here is the whole report on a Hacker News thread of 336 comments:

```{r}
#| echo: false
say(probe("corpus/02-hn-thread/source.json"))
```

Thirteen fields, one key-set, 336 copies, and **`RECURSIVE, 13 levels`**. A
comment thread has one record shape that contains itself, so describing it once
per level would be O(depth) for a structure whose entire point is that it
repeats. An earlier version printed the same thirteen fields twelve times and
then priced the thread at 25 rows, because it only ever saw the top level.

### The fold is a fixed point, not a pass

Data-ness is a property of the aggregate, and the aggregate only becomes visible
once the container above it has folded. One npm field is one to five keys per
copy and about thirty across copies, so a test applied per instance never folds
it and the output grows with the data. So the walk runs, folds whatever now
looks like data, and runs again, until nothing new folds.

### Reachability, or a document with no recursion reports some

Two paths having the same key set is not enough to call one a repeat of the
other. FHIR builds everything out of a handful of reusable element types, so
`Money` is `{value, currency}` in a dozen unrelated places. A document
containing **no recursion at all** was reported as recursive, twice.

The missing condition is **reachability**: you have to be able to get there by
following a field the ancestor actually has. A comment recurses because a
comment has a `children`. A `total` does not contain a `total` — the step into
`amount` leaves the shape behind and lands on something that merely looks the
same.

## Keys that are data

`{"1.0.0": {...}, "1.0.1": {...}}` is an object whose keys are values. Folding
288 of those into `<key>` is what makes the description finite — but folding
them *silently* would hide the thing that makes the file hard, so the fold is
always reported.

Deciding which is which takes two signals and **neither works alone**: how much
sibling copies share their key sets, and whether the values under one object
share a type. Alone, the first calls ragged records data; alone, the second
calls `author{name, email}` data.

There is a boundary no structural signal can cross, and the tool says so rather
than guessing. `{"text/html": …, "text/plain": …}` is data-as-keys with a
closed, stable vocabulary, which makes it structurally a record. Those are
reported **undecided**, and the honesty costs nothing:

```{r}
#| echo: false
r <- probe("corpus/01-npm-registry/source.json")
say(r[grep("KEYS THAT ARE DATA", r) : (grep("RECORD SHAPES", r) - 1)])
```

## The fold reporting that it should not have folded

One `entry` array in a FHIR bundle holds 564 resources of 20 different kinds.
Folded into one shape that is **97 fields, 87% empty, and exactly two fields
present in all of them.**

**The fold was never wrong. Its scope was.** "Sibling instances" was doing
unexamined work — siblings by position in a document are not siblings by kind —
and the document said so in the fold's own output, because 42 distinct key-sets
over 564 instances is the tool reporting that it merged things which are not the
same. Nothing read that number.

Now something does:

```{r}
#| echo: false
r <- probe("corpus/05-fhir-bundle/source.json")
i <- grep("SPLIT ON\\s+resourceType", r)[1]
say(r[(i - 1):(i + 7)])
```

A discriminator has to earn five guards, and every one of them is a false
positive that appeared while fitting this. It has to be present in every
instance, be a scalar, take few distinct values, at least halve the key-sets per
group — **and pay in holes**, meaning the worst group must be at most half as
empty as the fold was. That last one is the operation's own definition rather
than a threshold: the entire reason to split is that the folded table is mostly
holes, so a split that leaves the holes has not done the thing.

::: {.unsettled}
**The limit, and it is load-bearing.** The discriminator must live *inside* the
record. A GitHub event archive's payloads differ by event type, but `type` sits
on the parent event and not on the payload, so nothing here finds it. That is
the obvious next case and it is not solved.
:::
