---
title: "What makes a document hard"
---

```{r}
#| include: false
source("R/setup.R")
```

Naming what makes a document hard is most of the way to the syntax, so the axes
are a primary output of this project rather than a way of organising it. A file
is graded on each independently, because a file can be brutal on one and trivial
on the rest.

One rule governs the list:

> **An axis that never separates two files is not an axis.**

| Axis | What it means |
|---|---|
| **depth** | how many levels down the values live |
| **recursion** | self-similar structure of unknown depth, such as a comment thread |
| **raggedness, by absence** | a key present in some records and absent in others |
| **raggedness, by null** | the key is always there and the value sometimes is not |
| **polymorphism** | one field is a scalar here and an object there |
| **keys-as-data** | object keys are values, `{"1.0.0": {…}, "1.0.1": {…}}` |
| **heterogeneous arrays** | one array holding more than one shape |
| **path variance** | one field *name* living under more than one container |
| **unit ambiguity** | more than one defensible answer to "what is one row" |
| **scale** | does it fit in memory |

## The three corrections, which are the interesting part

**One axis was demoted.** *Path explosion* — the sheer number of distinct paths
— is high exactly when keys-as-data or recursion is present and low when neither
is. It carries no information its causes do not, so it is still reported,
because it is what you notice first, and no longer graded as independent.

**One axis was two axes.** *Raggedness* was doing two jobs. One corpus file is
ragged by absence and not at all by null; another is the exact reverse, with one
key-set across every node and four fields that are null on all but one of them.
Scored on the old single axis the second reads "raggedness: none", which is true
and thoroughly misleading.

**One axis is narrower than its own description.** *Path variance* was described
as "the same logical value living at different paths", which covers two things,
and only one of them was ever measured:

| | what it is | measured? |
|---|---|---|
| **by relocation** | one field NAME under several containers | **yes, this is the axis** |
| **by renaming** | one logical FIELD under several names — `Rating` and `rating` | **no axis measures it** |

**Renaming is recorded as a real property with no instrument**, rather than
folded into an axis that would then mean two things. The obvious detector —
fields under one container that never co-occur — was measured across the corpus
and catches *fields belonging to different kinds* instead: it pairs `Genre` with
`popcornscore`, which are mutually exclusive and are not the same field. What
makes `Rating` and `rating` one field is that the names resemble each other, and
this project refuses lexical rules.

**And one axis has never been tested.** No document in the corpus has ever
strained memory, so `scale` has never had the chance to separate anything. That
is a gap in the corpus, not evidence against the axis.

## Unit ambiguity is the one the tool is really for

The cost of rectangling **changes in kind with the row you pick.** Shallow gives
you holes; deep gives you duplication. The document tells you neither.

On the first corpus file, one row per version is 288 rows that are 60% empty,
and one row per dependency edge is 4,645 rows with the package name repeated
4,645 times. Both are correct answers to "what is one row".

```{r}
#| echo: false
r <- probe("corpus/01-npm-registry/source.json")
say(r[grep("ONE ROW COULD BE", r):length(r)])
```

Pricing those is the part nothing else offers, and it is [the third
question](rows.qmd).
