---
title: "Is this sound?"
---

```{r}
#| include: false
source("R/setup.R")
```

**There is no such thing as a broken data frame, and that asymmetry is what
makes this fathom's job.** A data frame is validated by construction: if it
exists, it is well formed. JSON is bytes claiming to be a format. So the
rectangular world has no verb for *is this even the thing it says it is*, and
nothing transfers.

This is also the one place fathom breaks its own rule about what earns a word. A
word that touches the data has to work at more than one depth; a health check
has no depth-one analogue **by construction**, and the very asymmetry it fails
is what justifies it.

## Loud and silent, not valid and invalid

The useful split is not the one a parser makes.

**Loud** damage announces itself. A truncated download. A model response cut off
at `max_tokens`, which is now the most common broken JSON there is. A JavaScript
object literal with unquoted keys. The comments and trailing commas of JSONC,
which is every `tsconfig.json`. NDJSON handed to a JSON parser.

**Silent** damage is the half worth owning, because the file parses and
something is lost anyway:

| | |
|---|---|
| duplicate keys | valid per the spec; the last one quietly wins |
| integers past 2^53 | anything JavaScript-derived rounds them |
| `NaN` and `Infinity` | Python's `json.dumps` emits them and jsonlite refuses them, so **a file Python wrote is a file R cannot read** |
| a negative zero | `json.loads("-0")` returns `0` and the sign is gone |
| an encoded document | a field whose value is itself a JSON document, as a string |

**The policy is report, never repair.** Silently fixing a document destroys the
evidence that something upstream is broken.

## What it looks like

Two lines, always in the same place, whatever the document:

```{r}
#| echo: false
say(head(probe("corpus/16-movie-ratings/source.json"), 5))
```

The same verb on a file that is **not one JSON document at all.** This one is
made up rather than drawn from the corpus, because every committed corpus file
is ordinary JSON and the point here is the format that is not:

```{r}
#| echo: false
nd <- file.path(tempdir(), "events.ndjson")
writeLines(c('{"id":1,"kind":"push"}', '{"id":2,"kind":"fork"}',
             '{bad line}',            '{"id":4,"kind":"push"}'), nd)
say(head(probe(nd), 6))
```

**Three records read, one line named, and the format still correctly called.**

## NDJSON is in scope, and the health check is what forces it

An NDJSON file is *not valid JSON*. A naive check calls it broken when it is
fine, so telling **broken** from **a different format** is unavoidable — and
NDJSON is how JSON at scale actually arrives: logs, warehouse exports, model
traces.

It stays NDJSON even when a line inside it is broken. An early version detected
the format, hit a bad line, and forgot — reporting "unrecognised" for a file
whose format it had already identified, which is the one case the format most
needs reporting for.

## The probe never lies about its own coverage

You cannot know whether a 10 GB file is chopped off without reading to the end,
and you cannot probe it by parsing it. So the report states how many records it
read and what it therefore cannot claim:

> `SAMPLE: the first 20,000 of 1,247,913 records. Everything below describes
> those and cannot speak for the rest.`

The same honesty runs deeper than it looks. When nothing could be parsed, the
report carries **no damage counters at all** rather than zeroes — because "0
duplicate keys" is a claim about a document it never got to look at. That
distinction survived the port to Rust as a deliberate piece of work.

## The one detector that cried wolf

The `NaN` check used to run a regular expression over the raw text. On a Jupyter
notebook it fired ten times, every one of them the word "NaN" inside legitimate
R console output that the notebook had stored as a string.

**A health check that cries wolf is worse than none**, because the next real one
gets ignored. The detector now counts parsed values, not text.

The same lesson arrived a second time from the other direction. A check for
"values that are themselves encoded JSON" reported 17 on the same notebook, and
every one was a false positive: Advent of Code day 18 is *Snailfish*, whose
puzzle input **is** nested integer lists, and `[376.0, 490.543]` is a Python
`repr`. All 17 parse as JSON and start with a bracket. The test was right by its
own definition and wrong about the world, and the fix was to say what a
*document* is: an object, or an array containing one. A bare array of scalars is
data that happens to be bracketed.

## How it is scored

Health is the one part of fathom with an exhaustive test suite rather than a
corpus, and the difference is a rule. The corpus takes real files only, because
toy JSON is hard in ways nobody suffers from. A test suite is the opposite:
synthetic, exhaustive, generated.

**What is scored is what fathom *said*, not whether it parsed.** A parser's
suite asks *did you accept it?* A reporter's asks *did you say the right thing
about it?* So a flag that fires when it should not fails exactly as loudly as
one that does not fire when it should.

The suite includes a **truncation ladder** — every single byte offset of three
documents, each one cut short — and a **control for every damage flag**, a case
that is genuinely clean and where the flag must stay silent. The controls are
there because of the `NaN` incident above.

```{r}
#| echo: false
say(tail(at_root(system2("uv", c("run", "test/check.py"),
                         stdout = TRUE, stderr = FALSE)), 9))
```

Those same cases score the Rust core, through the same harness. [Parity](parity.qmd)
is about why that had to be the *same* scorer rather than a second one.
