---
title: "Parity"
---

```{r}
#| include: false
source("R/setup.R")
```

A rewrite is the easiest place in software to lose something without noticing.
The new thing passes its tests, the old thing is deleted, and the behaviour that
nobody wrote a test for is gone.

This port is arranged so that cannot happen, and the arrangement is one
sentence:

> **`design/probe.py` is not replaced by the port. It becomes the oracle, and
> the port is finished when the difference between them is empty.**

## The port reproduces the defects too

That follows from the sentence above and is worth stating on its own, because
the temptation runs the other way.

The prototype has open defects. It cannot see a structure whose parts are
related by position rather than nesting when there is only one copy of the
holder. Its natural extraction yields list-columns that the sibling project's
spec refuses. Both are known, both are written down, and **both were ported
faithfully.**

A port that improves what it copies cannot be verified against it, because the
improvement and the regression arrive as the same diff. The repairs happen
afterwards, deliberately, one at a time, against an oracle that agrees.

## The stages say where, the page says whether

The criterion is a byte diff of the finished report, and nothing weaker will do.
A report that is right in substance and different in layout is still a different
report — and this project's own coverage instrument reads the page back as a set
of claims, telling a shape header from a split from a wrapped continuation **by
indentation alone**. A column is not cosmetic here.

But a byte diff alone can only say *this document differs*, and leave which of
the walk, the fold, the classifier, the pricing or the renderer produced it to
be guessed. So every stage is dumped and diffed as structured data as well.

**Both were necessary, and the last defect in the whole port proves it.** Two
documents differed at the very end, on twelve lines each. Every stage
comparison had already passed: the aligned-arrays detector had found the right
paths, the right lengths, the right parents and the right header candidates on
both files. **Every finding was correct and the page was still wrong**, because
one format string padded a path to 46 characters and then wrote the next column
without the space between them.

No structured comparison of the fold would ever have said so.

```{r}
#| echo: false
say(at_root(system2("uv", c("run", "test/parity.py"),
                    stdout = TRUE, stderr = FALSE)))
```

The health verb is compared on the synthetic cases **and** every corpus file,
field by field. The structure stages are compared on every document that parses.

## One scorer, two implementations

The health suite scores the Rust core through exactly the harness that scores
the Python. That was a deliberate choice over the obvious alternative of writing
a Rust test suite.

**A separate suite is free to be wrong in the same direction as the code it
scores**, which is how two implementations drift while both stay green. One
manifest and one set of rules read both, so a disagreement is a failure of
whichever is being scored and cannot hide in the harness.

## The check the pass/fail suite could not make

The 198 health cases record which damage flags should **fire**. They do not
record what those flags should **count**.

That gap matters, because the prediction written down before the port began was
that the *counts* would diverge — specifically that Python's UTF-8 replacement
and Rust's would disagree about how many replacement characters a truncated
document had earned. Both implementations could pass all 198 cases while
disagreeing about exactly the number the prediction was about.

So the field-by-field diff exists beside the pass/fail suite. **The prediction
turned out to be wrong** — 145 truncation cases agree byte for byte — but it was
wrong *on the record*, which is the only way a prediction is worth making.

## What the diff actually caught

One thing, in the whole port, and it is a fact about CPython rather than about
Rust.

Its scanner wants **five** characters after a `\u` escape, not four. So a
document truncated exactly after the four hex digits reports `Invalid \uXXXX
escape` where every defensible reading says `Unterminated string`. Five cases in
the truncation ladder land on precisely that byte.

**Its own source documents the guard as `> len` and it behaves as `>=`.** No
amount of reading the C would have found it; only running both and diffing did.

The port copies the behaviour, not the source, because the oracle is CPython.

## Two things reproduced that are not in the probe's source

**Python's `set` iteration order decides a tie.** The recursion fold picks a
canonical path out of a set, and set order is unspecified and hash-randomised.
Rather than assume it never matters, every corpus file was rendered under
randomised hash seeds and checked: all identical. The port uses a deterministic
order, and the sweep is what makes that safe rather than lucky.

**`True == 1` in Python**, so a discriminator grouping on raw values treats them
as one kind. No corpus file exercises it. It is reproduced anyway, because a
divergence there would be invisible.
