---
title: "The fixed questions"
---

```{r}
#| include: false
source("R/setup.R")
```

Every file is asked the same questions, in every tool, or nothing is comparable.
The list is `QUESTIONS.md` in the repository, and it is the specification for
the vocabulary as much as it is the method:

> **A word belongs in fathom only if removing it makes one of the fixed
> questions unanswerable on at least one corpus file. No word is added without
> naming the question it answers and the file that proves it.**

That makes the question list the vocabulary's bound. It also explains why one
existing library has roughly 180 exports and fathom expects to have a handful:
that library answers questions nobody in this corpus asked.

```{r}
#| echo: false
q <- readLines(file.path(ROOT, "QUESTIONS.md"))
say(grep("^\\s*[0-9]+\\.", q, value = TRUE))
```

## Two rules that make the answers mean something

**Score the exploring separately from the extracting.** The whole premise is
that exploring an unknown document is where the cost is and that nobody measures
it. A session that records only how the extraction went has measured the half
that was never in question. So for each file and tool: how long until you could
*state* the shape, and what you had to run to get there.

**A question added after the feature it describes is marked as such.** Scoring
another tool "cannot" on a question written to match something fathom had just
decided to do is circular, and a reader needs to be able to tell those apart
from the questions that predate the design.

## "Cannot" is the most useful cell in the grid

A tool that cannot answer a question records that, with the reason, and moves
on. An empty cell is the least useful thing an attempt file can contain, because
it does not distinguish *the tool refused* from *nobody tried*.
