---
title: "What could one row be?"
---

```{r}
#| include: false
source("R/setup.R")
```

**This half is ported and verified.** The Rust core prices candidates
identically to the prototype on every corpus file — including reproducing
`pandas.json_normalize`, which was named as the port's largest risk before any
Rust was written. [Parity](parity.qmd) has that story.

**Rectangling is not the goal**, and getting that wrong was the first
correction this project made to itself.

> **A table is the shape of your answer, never the shape of the document.**

An extract may be any shape — a value, a vector, a list. The rectangular ones
are what flow onward into the next tool in the chain; non-rectangular extracts
are terminal, for looking at or for feeding back into another call.

But when you *do* want a table, the document does not tell you which one, and
**pricing the candidates is the part nothing else offers.**

## Every defensible answer, and what each costs

```{r}
#| echo: false
r <- probe("corpus/01-npm-registry/source.json")
say(r[grep("ONE ROW COULD BE", r):length(r)])
```

Read the two extremes. One row per version is a wide table that is mostly holes.
One row per dependency edge is a long table with the same name repeated
thousands of times. **The cost changes in kind with the row you pick** — shallow
gives you holes, deep gives you duplication — and neither is wrong.

## Computed on the fold, never on the data

The first version of this recursed into raw values and emitted one candidate per
version, producing **1,239 lines on a 786 KB file** — reproducing the exact
failure the whole project exists to prevent.

> **The fold is not a display step.** Anything computed from raw values rather
> than from the folded structure will be proportional to the data no matter what
> it prints.

## The candidate that is really several tables

A document of ten kinds has ten answers to "what is one row", and the operation
that finds them has already run by the time the candidates are priced. An
earlier version printed the split near the top of the report and then, twenty
lines later, offered a single 319-column table that was 93% empty — same
program, same fold, two numbers, no join.

```{r}
#| echo: false
r <- probe("corpus/02-hn-thread/source.json")
say(r[grep("ONE ROW COULD BE", r):length(r)])
```

The `└─` line is the report contradicting its own previous line, on purpose.

## What porting this costs

The prototype prices a candidate by building the table with a data-frame library
and reading three things off it: its shape, its fraction of missing cells, and
the most-repeated value in each column.

**That was the single largest parity risk in the port**, named as such before
any Rust was written. The core has to reproduce that library's flattening rule
and its missing-value semantics exactly, or every percentage in the block above
moves — and a percentage that moves by one point is a diff that looks like a bug
in the fold.

It reproduces them, and two of the behaviours had to be **measured** because the
obvious guess was wrong. `astype(str)` leaves a missing value missing rather
than rendering it as the string `"nan"`, so the distinct count drops it. And the
inferred column type decides the digits: a column of integers with no hole
prints `1`, and the same column with one hole prints `1.0`. **Same data,
different count, and nothing in the document changed.**

::: {.unsettled}
**The cost is a standing liability.** That part of the core is a model of one
library's behaviour at one version. Anything pandas changes there is a parity
failure rather than a bug in fathom, and the harness is what would catch it.
:::
