---
title: "Four words, two languages"
---

```{r}
#| include: false
source("R/setup.R")
```

::: {.unsettled}
**None of this is a package yet, and the distinction matters more here than
anywhere else in the book.** There are now **three** implementations of these
four words and not one of them ships: the Python prototypes in `design/`, a
hand-written R re-implementation, and the Rust core. What is missing is the
thin R and Python packages that would call the core — and until those exist,
nobody can install any of it.

The three agree on every sentence in this chapter. **That is a fact about the
plumbing and not about the vocabulary**, which is the distinction the end of
this chapter is about.
:::

## `rows` — say what a row is, not where it lives

The hypothesis in one line: every existing tool makes you state **paths**, from
which the row shape emerges as a side effect. `rows` makes you state what a row
is.

::: {.panel-tabset}

## R

```r
rows(doc, "versions.*")
```

## Python

```python
rows(doc, "versions.*")
```

:::

`*` is "every key here, whatever it is called" — which is what makes it work on
a document whose keys are data. The keys it matched come back beside the
records, because on a keys-as-data document **the key is the most important
column** and a tool that discards it has thrown away the version number.

## `first_present` — a priority order, not a set

The first of these paths that is actually there.

::: {.panel-tabset}

## R

```r
first_present(node, "valueQuantity", "valueString", "valueCodeableConcept",
              default = NULL)
```

## Python

```python
first_present(node, "valueQuantity", "valueString", "valueCodeableConcept",
              default=None)
```

:::

**Both halves of the name carry something people get wrong.** `first` says the
arguments are a priority order rather than a set. `present` says the only value
skipped is a missing one — **so a zero comes back.**

This is the first word shared with the sibling project, where it is the same
idea over a table rather than a document. Sharing a word across two projects
needs a stated owner and a test on both sides from the day it is shared, or the
two drift and the drift is invisible.

## `take` — one row per record, one column per path

::: {.panel-tabset}

## R

```r
take(records, "name", "author.email", "dist.tarball")
```

## Python

```python
take(records, "name", "author.email", "dist.tarball")
```

:::

**Nothing else is built**, and that is the whole point. The measured cost claim
is that building only the columns you asked for skips **97.9%** of the cells a
general-purpose flattener would materialise.

## `where` — find the value, not the path

The verb for a document you have never seen: every path whose value matches,
**folded**.

::: {.panel-tabset}

## R

```r
where(doc, "url")
```

## Python

```python
where(doc, "url")
```

:::

Folded, because otherwise it is the O(data) failure arriving through the front
door — 5,511 matching paths in one corpus file, which is a wall of text rather
than an answer. It reports the **shapes** instead.

## The claim, checked

The two implementations were written independently and are diffed on real
documents. This is the harness running now:

```{r}
#| echo: false
say(at_root(system2("uv", c("run", "design/parity.py"),
                    stdout = TRUE, stderr = FALSE)))
```

## Why a second implementation was written at all

This looks like exactly the mistake the architecture forbids. **One Rust core,
so that two hand-written copies can never drift** — and here are two
hand-written copies.

The distinction is that one of them is a probe rather than a product, and it
answers a question the core **cannot**:

> One core makes the two languages agree by construction. That proves the
> plumbing. **It cannot tell you whether the vocabulary is right.** If
> `rows("versions.*")` reads naturally in Python and awkwardly in R, a shared
> core delivers the awkward reading to R users faithfully and forever. The port
> does not test that assumption, it locks it in.

The only way to test it is to have the words re-derived from the written
notation by another route, and see whether they land in the same place. They
did, on every sentence above.

**Two hand-written copies must never be *shipped*, because they drift
invisibly. Writing one as a probe and throwing it away costs nothing and buys
the one piece of evidence the architecture assumes.**

### The core now answers these sentences, and the R copy has not been deleted

The plan said `design/fathom.R` goes the day the core answers them. It does —
`rows` and `where` in Rust land on the same figures as both the Python and the
R, on all nineteen.

**It is kept anyway, and the reason is worth more than the rule.** The same plan
keeps the parity harness by *retargeting* it from two implementations to two
bindings of one core. There are no bindings yet. Deleting the R today would
break that harness and leave the core checked against one implementation where
it is currently checked against two — trading a working check for nothing, to
be tidy.

It goes when the R package exists to replace it. Written down as a deliberate
deferral, because a plan followed to the letter at the cost of the thing it was
protecting is not being followed at all.
